---
title: "Introduction to MeatColor"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to MeatColor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```
```{r setup}
library(MeatColor)
library(dplyr)
library(ggplot2)
```

## Overview

The `MeatColor` package provides tools to visualize L\*a\*b\* colorimeter data commonly used in meat science research. The package includes two main functions:

- `summarize_lab()`: Summarizes L\*a\*b\* color measurements by grouping variables
- `plot_lab_colors()`: Creates visual representations of color data as colored rectangles

## Basic Workflow

### Step 1: Prepare Your Data

Your data should contain:
- L\*, a\*, b\* color measurements (typically named `l`, `a`, `b`)
- Grouping variables (e.g., treatment, time point, product)
```{r example-data}
# Example dataset
color_data <- data.frame(
  product = rep(c("Beef", "Pork"), each = 12),
  treatment = rep(rep(c("Control", "Vacuum"), each = 6), times = 2),
  time_point = rep(rep(c("Day 0", "Day 7", "Day 14"), each = 2), times = 4),
  replicate = rep(1:2, times = 12),
  l = c(45.2, 44.8, 42.1, 41.9, 39.5, 39.2,
        46.1, 45.9, 43.2, 42.8, 40.1, 39.9,
        48.3, 48.1, 45.5, 45.2, 42.8, 42.5,
        49.2, 48.9, 46.3, 46.1, 43.5, 43.2),
  a = c(18.5, 18.8, 16.2, 16.5, 14.1, 14.3,
        17.9, 18.1, 15.8, 16.1, 13.9, 14.2,
        12.3, 12.5, 10.8, 11.1, 9.5, 9.7,
        11.8, 12.1, 10.5, 10.7, 9.2, 9.4),
  b = c(12.1, 12.3, 10.5, 10.7, 9.2, 9.4,
        11.8, 12.0, 10.3, 10.5, 9.0, 9.2,
        8.5, 8.7, 7.2, 7.4, 6.1, 6.3,
        8.2, 8.4, 7.0, 7.2, 5.9, 6.1)
)

head(color_data)
```

### Step 2: Summarize the Data

Use `summarize_lab()` to calculate mean L\*a\*b\* values for each combination of grouping variables:
```{r summarize}
color_summary <- summarize_lab(
  data = color_data,
  group_vars = c("product", "treatment", "time_point")
)

color_summary
```

The function automatically:
- Calculates mean L\*, a\*, b\* values
- Converts them to hex colors for visualization
- Handles missing values with `na.rm = TRUE`

### Step 3: Visualize the Colors

Create a color plot showing how colors change across conditions:
```{r plot-basic, fig.width=8, fig.height=4}
plot_lab_colors(
  data = color_summary,
  x_var = "time_point",
  facet_var = "product",
  x_label = "Storage Duration",
  title = "Meat Color Changes During Storage"
)
```

## Advanced Usage

### Custom Column Names

If your L\*a\*b\* columns have different names:
```{r custom-cols, eval=FALSE}
# Data with columns named L_star, a_star, b_star
summarize_lab(
  data = my_data,
  group_vars = c("treatment", "day"),
  l_col = "L_star",
  a_col = "a_star",
  b_col = "b_star"
)
```

### Single Grouping Variable
```{r single-group}
# Summarize by time point only
time_summary <- summarize_lab(
  data = color_data,
  group_vars = "time_point"
)

plot_lab_colors(
  data = time_summary,
  x_var = "time_point",
  x_label = "Day",
  title = "Overall Color Change"
)
```

### No Faceting
```{r no-facet}
# Plot without faceting
treatment_summary <- summarize_lab(
  data = color_data %>% filter(product == "Beef"),
  group_vars = c("treatment", "time_point")
)

plot_lab_colors(
  data = treatment_summary,
  x_var = "time_point",
  x_label = "Day"
)
```

### Customizing Plots

Since `plot_lab_colors()` returns a ggplot object, you can add additional layers:
```{r customize, fig.width=8, fig.height=4}
plot_lab_colors(
  data = color_summary,
  x_var = "time_point",
  facet_var = "treatment"
) +
  labs(caption = "Data from XYZ study, 2024") +
  theme(
    plot.caption = element_text(hjust = 0, size = 10),
    strip.background = element_rect(fill = "lightblue")
  )
```

## Interpreting Results

The colored rectangles represent the actual color as determined by the L\*a\*b\* values:

- **L\*** (lightness): 0 = black, 100 = white
- **a\***: negative = green, positive = red
- **b\***: negative = blue, positive = yellow

This visualization allows you to quickly see:
- Color changes over time
- Differences between treatments
- Product-specific color characteristics

## Tips

1. **Ensure consistent grouping**: All grouping variables should be factors or characters
2. **Handle missing data**: The functions use `na.rm = TRUE` by default
3. **Check your data**: Make sure L\*, a\*, b\* values are in typical ranges (L\*: 0-100, a\* and b\*: typically -60 to +60)
4. **Export plots**: Use `ggsave()` to save plots for publications
```{r save-plot, eval=FALSE}
p <- plot_lab_colors(color_summary, x_var = "time_point", facet_var = "product")
ggsave("color_plot.png", p, width = 8, height = 4, dpi = 300)
```
