---
title: "Introduction to MeatColor"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to MeatColor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```
```{r setup}
library(MeatColor)
library(dplyr)
library(ggplot2)
```

## Overview

The `MeatColor` package provides tools to summarize, compare, and visualize L\*a\*b\* colorimeter data commonly used in meat science research. The package includes nine main functions:

- `summarize_lab()`: Summarizes L\*a\*b\* color measurements by grouping variables
- `plot_lab_colors()`: Creates visual representations of color data as colored rectangles
- `lab_to_hex()`: Converts CIELAB coordinates to approximate display colors
- `lab_chroma()`: Calculates CIELAB chroma from a\* and b\* coordinates
- `lab_hue()`: Calculates CIELAB hue angles from a\* and b\* coordinates
- `add_lab_metrics()`: Adds hue and chroma columns to a data frame or tibble
- `delta_e_2000()`: Calculates perceptual color differences using CIEDE2000
- `myoglobin_int()`: Estimates OMb, DMb, and MMb from MiniScan reflectance
- `myoglobin_ref()`: Uses calibrated OMb, DMb, and MMb reference spectra
- `myoglobin_calibration()`: Creates a reusable calibrated reference object

## Basic Workflow

### Step 1: Prepare Your Data

Your data should contain:
- L\*, a\*, b\* color measurements (typically named `l`, `a`, `b`)
- Grouping variables (e.g., treatment, time point, product)
```{r example-data}
# Example dataset
color_data <- data.frame(
  product = rep(c("Beef", "Pork"), each = 12),
  treatment = rep(rep(c("Control", "Vacuum"), each = 6), times = 2),
  time_point = rep(rep(c("Day 0", "Day 7", "Day 14"), each = 2), times = 4),
  replicate = rep(1:2, times = 12),
  l = c(45.2, 44.8, 42.1, 41.9, 39.5, 39.2,
        46.1, 45.9, 43.2, 42.8, 40.1, 39.9,
        48.3, 48.1, 45.5, 45.2, 42.8, 42.5,
        49.2, 48.9, 46.3, 46.1, 43.5, 43.2),
  a = c(18.5, 18.8, 16.2, 16.5, 14.1, 14.3,
        17.9, 18.1, 15.8, 16.1, 13.9, 14.2,
        12.3, 12.5, 10.8, 11.1, 9.5, 9.7,
        11.8, 12.1, 10.5, 10.7, 9.2, 9.4),
  b = c(12.1, 12.3, 10.5, 10.7, 9.2, 9.4,
        11.8, 12.0, 10.3, 10.5, 9.0, 9.2,
        8.5, 8.7, 7.2, 7.4, 6.1, 6.3,
        8.2, 8.4, 7.0, 7.2, 5.9, 6.1)
)

head(color_data)
```

### Step 2: Summarize the Data

Use `summarize_lab()` to calculate mean L\*a\*b\* values for each combination of grouping variables:
```{r summarize}
color_summary <- summarize_lab(
  data = color_data,
  group_vars = c("product", "treatment", "time_point")
)

color_summary
```

The function automatically:
- Calculates mean L\*, a\*, b\* values
- Converts them to hex colors for visualization
- Handles missing values with `na.rm = TRUE`

### Step 3: Visualize the Colors

Create a color plot showing how colors change across conditions:
```{r plot-basic, fig.width=8, fig.height=4}
plot_lab_colors(
  data = color_summary,
  x_var = "time_point",
  facet_var = "product",
  group_var = "treatment",
  group_label = "Treatment",
  x_label = "Storage Duration",
  title = "Meat Color Changes During Storage"
)
```

## Advanced Usage

### Custom Column Names

If your L\*a\*b\* columns have different names:
```{r custom-cols, eval=FALSE}
# Data with columns named L_star, a_star, b_star
summarize_lab(
  data = my_data,
  group_vars = c("treatment", "day"),
  l_col = "L_star",
  a_col = "a_star",
  b_col = "b_star"
)
```

### Single Grouping Variable
```{r single-group}
# Summarize by time point only
time_summary <- summarize_lab(
  data = color_data,
  group_vars = "time_point"
)

plot_lab_colors(
  data = time_summary,
  x_var = "time_point",
  x_label = "Day",
  title = "Overall Color Change"
)
```

### No Faceting
```{r no-facet}
# Plot without faceting
treatment_summary <- summarize_lab(
  data = color_data %>% filter(product == "Beef"),
  group_vars = c("treatment", "time_point")
)

plot_lab_colors(
  data = treatment_summary,
  x_var = "time_point",
  group_var = "treatment",
  x_label = "Day"
)
```

### Customizing Plots

Since `plot_lab_colors()` returns a ggplot object, you can add additional layers:
```{r customize, fig.width=8, fig.height=4}
plot_lab_colors(
  data = color_summary,
  x_var = "time_point",
  facet_var = "product",
  group_var = "treatment",
  subtitle = "Instrumental color by product and treatment",
  caption = "Rectangles show group means",
  base_size = 10,
  border_linewidth = 0.4,
  show_values = "hex"
) +
  labs(caption = "Data from XYZ study, 2024") +
  theme(
    plot.caption = element_text(hjust = 0, size = 10),
    strip.background = element_rect(fill = "lightblue")
  )
```

### Calculating Hue and Chroma

Chroma measures distance from the neutral axis, while hue angle describes the
direction around the a\*–b\* color plane. Both functions are vectorized:

```{r hue-chroma}
lab_chroma(a = c(18, 12), b = c(12, 8))
lab_hue(a = c(18, 12), b = c(12, 8))
```

Use `add_lab_metrics()` to add both values to a data frame and continue directly
into grouped statistics:

```{r dataframe-hue-chroma}
metric_summary <- color_data |>
  add_lab_metrics(a_col = a, b_col = b) |>
  group_by(product, treatment) |>
  summarize(
    mean_chroma = mean(chroma, na.rm = TRUE),
    mean_hue = mean(hue, na.rm = TRUE),
    .groups = "drop"
  )

metric_summary
```

`lab_hue()` returns degrees by default: 0° is red, 90° is yellow, 180° is
green, and 270° is blue. Use `degrees = FALSE` for radians. Neutral colors have
zero chroma and an undefined (`NA`) hue.

### Comparing Colors with CIEDE2000

Use `lab_distances()` to compare all sample combinations when the experiment
does not have a reference color. The resulting S3 object preserves the sample
metadata and can be converted to a numeric matrix, summarized by treatment, or
plotted directly.

```{r delta-e}
comparison_data <- data.frame(
  sample_id = paste0("S", 1:6),
  treatment = rep(c("Control", "Treated"), each = 3),
  l = c(45, 46, 44, 50, 51, 49),
  a = c(18, 17, 19, 13, 12, 14),
  b = c(12, 11, 13, 9, 8, 10)
)

distances <- lab_distances(comparison_data, sample_id = sample_id)
as.matrix(distances)
summarize_treatment_distances(distances, treatment = treatment)
plot(distances, treatment = treatment)
```

Treatment summaries use each unique pair once, exclude self-comparisons, and
report the number of complete and missing pairs. `delta_e_2000()` remains the
lower-level function for intentionally paired colors. The optional `k_l`,
`k_c`, and `k_h` arguments control the CIEDE2000 lightness, chroma, and hue
parametric weighting factors; their defaults are the reference-condition value
of 1.

### Estimating Myoglobin Redox Forms

`myoglobin_int()` accepts MiniScan reflectance measurements at 10 nm intervals
and appends estimated oxymyoglobin, deoxymyoglobin, and metmyoglobin
percentages. Specify the reflectance scale explicitly:

```{r myoglobin}
miniscan_data <- data.frame(
  sample = "A",
  R470 = 53.1206, R480 = 53.1206,
  R520 = 50.1187, R530 = 50.1187,
  R570 = 45.8142, R580 = 45.8142,
  R700 = 79.4328
)

myoglobin_int(miniscan_data, reflectance_scale = "percent")
```

The function interpolates reflectance at 473, 525, and 572 nm before converting
the values to reflex attenuance. It uses 700 nm as the reference wavelength
because that is the upper end of the MiniScan spectral range. Estimates outside
0–100% are retained with a warning by default so potentially problematic scans
remain visible to the analyst.

When prepared 100% reference samples are available, create a reusable S3
calibration object for the calibrated AMSA K/S method:

```{r myoglobin-reference, eval=FALSE}
calibration <- myoglobin_calibration(
  omb_reference = omb_reference_scans,
  dmb_reference = dmb_reference_scans,
  mmb_reference = mmb_reference_scans,
  reflectance_scale = "percent"
)

reference_result <- predict(calibration, newdata = miniscan_data)
myoglobin_calibration_ratios(calibration)
plot(calibration)
```

All four data frames must contain the same 470, 480, 520, 530, 570, 580, and
610 nm columns. Replicate reference scans are summarized after their K/S ratios
are calculated. The OMb, DMb, and MMb estimates are calibrated independently,
so their sum is reported but is not forced to equal 100%. The calibration plot
is a regular ggplot object. `myoglobin_ref()` remains available as a concise
one-call interface when the references will only be applied once.

## Interpreting Results

The colored rectangles represent the actual color as determined by the L\*a\*b\* values:

- **L\*** (lightness): 0 = black, 100 = white
- **a\***: negative = green, positive = red
- **b\***: negative = blue, positive = yellow

This visualization allows you to quickly see:
- Color changes over time
- Differences between treatments
- Product-specific color characteristics

## Tips

1. **Ensure consistent grouping**: All grouping variables should be factors or characters
2. **Handle missing data**: The functions use `na.rm = TRUE` by default
3. **Check your data**: Make sure L\*, a\*, b\* values are in typical ranges (L\*: 0-100, a\* and b\*: typically -60 to +60)
4. **Export plots**: Use `ggsave()` to save plots for publications
```{r save-plot, eval=FALSE}
p <- plot_lab_colors(
  color_summary,
  x_var = "time_point",
  facet_var = "product",
  group_var = "treatment"
)
ggsave("color_plot.png", p, width = 8, height = 4, dpi = 300)
```
